The Precambrian paleogeography of Laurentia

### 2020-09-17 Manuscript/Laurentia_Paleogeo_Manuscript.tex
move manuscript into new latex template

\documentclass[11pt,letterpaper]{article}
\setcounter{section}{3}
\usepackage{threeparttable}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{pdfpages}
\usepackage{url}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[outercaption]{sidecap}
\doublespacing

\raggedright
\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0.0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.1 in
\parindent = 0.2 in

\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

\begin{document}

\begin{flushleft}
\section{The Precambrian paleogeography of Laurentia}

Nicholas L. Swanson-Hysell\textsuperscript{1}

\bigskip

\textsuperscript{1} Department of Earth and Planetary Science, University of California, Berkeley, CA 94720 USA

\end{flushleft}

\noindent\textit{This chapter is in preparation for the book Ancient Supercontinents and the Paleogeography of the Earth}

%\linenumbers

\subsection{Abstract}

Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its ca. 1.8 Ga amalgamation. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.

\subsection{Introduction and broad tectonic history}

Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}). Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; prov. -- province). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}}
\label{fig:tectonic_history}
\end{figure}

@@

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989a} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
\caption{\small{\textbf{Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard). Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}}}
\label{fig:Laurentia_poles}
\end{figure}

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021b}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2019a} are also included.

@@

%The poles on this figure should be labeled
\begin{figure}
\centering
\includegraphics[width=6 in]{../Figures/Superior_Slave_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces. The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of narrower Manikewan with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}}}
\label{fig:Superior_Slave_recons}
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
\label{fig:Laurentia_paleolatitude}
\end{figure}

@@

Sedimentary rocks whose deposition is associated with specific climatic conditions have the potential to provide insight into paleolatitude. Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation. Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}. Evaporite deposits are particularly compelling as paleolatitude constraints given that their deposition is interpreted to be associated with arid regions resulting from large-scale Hadley cell downwelling \citep{Evans2006a}. While moisture in the subtropics can change along with Earth's climate, the overall pattern of $\sim$10-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}. Using a compilation of paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones. This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale convection circulation cells.

Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
\begin{itemize}
%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}. The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited 
\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preservedwithin secondary silica \citep{White1984a}. The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}. Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}. Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}. There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.  These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 within the Helena Formation \citep{Evans2000c}. 
\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.  These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} . These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca. 1005 Ma detrital zircons \citep{Turner2016a}. The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca. 820 Ma \citep{Turner2016a}. These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}. The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}. The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
\end{itemize}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_evaporite_figure.pdf}
\caption{\small{\textbf{Left panel: The paleolatitude of evaporite deposits following the amalgamation of Laurentia as reconstructed by the simple Laurentia model shown in Fig. \ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic. Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}. The evaporite lines extend from the maximum to minimum age constraints while the points are at the preferred depositional age.  Right panel: The present-day location of evaporites reconstructed to the left. 1: Altyn Formation (Belt Supergroup), 2: Wallace/Helena Formations (Belt Supergroup), 3: Iqqittuq Formation (Bylot Supergroup), 4: Ten Stone Formation (Mackenzie Mountains Supergroup), 5: Minto Inlet Formation (Shaler Supergroup), 6: Redstone River Formation (Mackenzie Mountains Supergroup), 7: Kilian Formation (Shaler Supergroup), 8: Carboniferous Canadian Maritime, 9: Carboniferous Sverdrup, 10: Permian Midcontinental USA.}}}
\label{fig:Laurentia_evaporites}
\end{figure}

In Figure \ref{fig:Laurentia_evaporites}, the paleolatitude of these evaporite deposits are reconstructed using the

@@

Developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types. Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles. Paleomagnetic pole compilations themselves are evolving with better data and improved geochronology. Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress in Precambrian paleogeography.

>>>>

%% arara directives
% arara: xelatex
% arara: bibtex
% arara: xelatex
% arara: xelatex

%\documentclass{article} % One-column default
\documentclass[twocolumn, switch]{article} % Method A for two-column formatting

\usepackage{preprint}

%% Math packages
%\usepackage{amsmath, amsthm, amssymb, amsfonts}

%% Bibliography options
%\usepackage[numbers,square]{natbib}
%\bibliographystyle{unsrtnat}
\usepackage{natbib}
%\bibliographystyle{Geology}

%% General packages
\usepackage[utf8]{inputenc}	% allow utf-8 input
\usepackage[T1]{fontenc}	% use 8-bit T1 fonts
\usepackage{xcolor}		% colors for hyperlinks
\usepackage[colorlinks = true,
            linkcolor = purple,
            urlcolor  = blue,
            citecolor = black,
            anchorcolor = black]{hyperref}	% Color links to references, figures, etc.
\usepackage{booktabs} 		% professional-quality tables
\usepackage{nicefrac}		% compact symbols for 1/2, etc.
%\usepackage{microtype}		% microtypography
%\usepackage{lineno}		% Line numbers
\usepackage{float}			% Allows for figures within multicol
\usepackage{textcomp,marvosym}
%\usepackage{multicol}		% Multiple columns (Method B)

%\usepackage{lipsum}		%  Filler text
\usepackage{lettrine}
\usepackage{pdflscape}
\usepackage{longtable}

 %% Special figure caption options
\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Supplementary Figure}]{suppfigure}
\usepackage{sidecap}
\sidecaptionvpos{figure}{c}

% Section title spacing  options
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 3pt minus 3pt}{1pt plus 1pt minus 1pt}
\titlespacing\subsection{0pt}{10pt plus 3pt minus 3pt}{1pt plus 1pt minus 1pt}
\titlespacing\subsubsection{0pt}{8pt plus 3pt minus 3pt}{1pt plus 1pt minus 1pt}

%%%%%%%%%%%%%%%%   Title   %%%%%%%%%%%%%%%%
\title{The Precambrian paleogeography of Laurentia}

% Add watermark with submission status
%\usepackage{xwatermark}
%% Left watermark
%\newwatermark[firstpage,color=gray!60,angle=90,scale=0.32, xpos=-4.05in,ypos=0]{\href{https://doi.org/}{\color{gray}{Publication doi}}}
%% Right watermark
%\newwatermark[firstpage,color=gray!60,angle=90,scale=0.32, xpos=3.9in,ypos=0]{\href{https://doi.org/}{\color{gray}{Preprint doi}}}
% Bottom watermark
%\newwatermark[firstpage,color=gray!90,angle=0,scale=0.28, xpos=0in,ypos=-5in]{*correspondence: \texttt{swanson-hysell@berkeley.edu}}

%%%%%%%%%%%%%%%  Author list  %%%%%%%%%%%%%%%
\usepackage{authblk}
\renewcommand*{\Authfont}{\bfseries}
\author[]{Nicholas L. Swanson-Hysell}
\affil[]{Department of Earth and Planetary Science, University of California, Berkeley, CA 94720 USA}

\setcounter{section}{4}

%%%%%%%%%%%%%%    Front matter    %%%%%%%%%%%%%%
\begin{document}

\twocolumn[ % Method A for two-column formatting
  \begin{@twocolumnfalse} % Method A for two-column formatting

\maketitle

\begin{abstract}
Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its ca. 1.8 Ga amalgamation. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years. 
\end{abstract}
%\keywords{First keyword \and Second keyword \and More} % (optional)
\textit{This manuscript is a preprint of the chapter: \vspace{0.1 cm} \\
Swanson-Hysell, N. L. (2021) Chapter 4: The Precambrian paleogeography of Laurentia. In: Pesonen, L.J., Salminen, J., Evans, D.A.D., Elming, S.-Å., Veikkolainen, T. (eds.) Ancient Supercontinents and the Paleogeography of the Earth.}
\vspace{0.3 cm}
  \end{@twocolumnfalse} % Method A for two-column formatting
] % Method A for two-column formatting

%\begin{multicols}{2} % Method B for two-column formatting (doesn't play well with line numbers), comment out if using method A

%%%%%%%%%%%%%%%  Main text   %%%%%%%%%%%%%%%
% \linenumbers

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989b, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989b}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\textbf{Simplified map of the tectonic units of Laurentia.} The location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust are simplified from \cite{Whitmeyer2007a} with additions for Greenland based on \cite{St-Onge2009a}. Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; prov. -- province). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel \citep{Evans2021a}.}
\label{fig:Laurentia_map}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}
\label{fig:tectonic_history}
\end{figure*}

@@

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989b} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Top panel: Paleomagnetic poles from 1800 to 560 Ma for Laurentia (including Greenland, Scotland and Svalbard). Bottom panel: Paleomagnetic data for Archean Provinces prior to the amalgamation of Laurentia.}
\label{fig:Laurentia_poles}
\end{figure*}

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021a}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2019a} are also included.

@@

%The poles on this figure should be labeled
\begin{figure*}
\centering
\includegraphics[width=6 in]{../Figures/Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of narrower Manikewan with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

\begin{table}[hbt]
\caption{Rotations of separated terranes}
{\scriptsize
\begin{tabular}{|l|l|l|l|p{1.1 in}|}
  \hline
& Euler pole & Euler pole & rotation & note and \\
Block & longitude & latitude & angle & citation \\
\hline
Greenland & -118.5 & 67.5 & -13.8 & Cenozoic separation of Greenland from Laurentia associated with opening of Baffin Bay and the Labrador Sea \citep{Roest1989a} \\
\hline
Scotland & 161.9 & 78.6 & -31.0 & Reconstructing Atlantic opening following \cite{Torsvik2017a} \\
\hline
Svalbard & 125.0 & -81.0 & 68 & Rotate Svalbard to Laurentia in fit that works well with East Greenland basin according to \cite{Maloof2006a}\\
\hline
\end{tabular}
}
\label{tab:terrane_rotations}
\end{table}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}
\label{fig:Laurentia_reconstructions}
\end{figure*}

@@

Developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types. Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles. Paleomagnetic pole compilations themselves are evolving with better data and improved geochronology \citep{Evans2021a}. Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress in Precambrian paleogeography.

There is an overall lack of models in the literature for the Proterozoic with published continuous rotation parameters that can be compared to the compilation of paleomagnetic poles presented herein. The approach in the community for many years has been to publish models as snapshots at given time intervals presented in figures without publishing continuous rotation parameters, although some studies have published the Euler rotations associated with specified times. With the further adoption of software tools such as GPlates, there has been significant progress in the publication of continuous paleogeographic models constrained by paleomagnetic poles through the Phanerozoic (540 Ma to present; e.g. \citealp{Torsvik2017a}).
